---
title: "Differential expression methylome with limma"
author: "Menna Arafat"
date:  "2024-9-02"
output: html_document
---

## create output directory

```{r}
dir.create("output")
```

## load libraries

```{r}
 suppressPackageStartupMessages({ 
library(limma)
library(tibble)
library(plyr)
library(dplyr)
library(tidyverse)
library("pheatmap")
library("dichromat")
library("RColorBrewer")
library(ComplexHeatmap)
library(circlize)   
 })
```

## load data

```{r}
data= read.csv("data/methyl_b_value.csv")

#filtration process if necessary
# count= data[1:100000,] %>% filter(., !is.na(gene_symbol) ) %>% 
#                           filter(., !duplicated(gene_symbol) ) %>% 
#                           filter(., rowSums(is.na(.)) < ncol(data)/2 ) %>% 
#                           column_to_rownames("gene_symbol") %>% 
#                           dplyr::select(-1)

#colnames(count)= c(paste0("sample_", 1:110, "_Tumor"), paste0("sample_", 111:ncol(count), "_Normal"))
```

## impute missing values

```{r}
count= data %>% as.data.frame() %>% column_to_rownames("X")
count_T = t(apply(count[1:110], 1, function(x) {
                  x[is.na(x)] <- median(x, na.rm = TRUE)
                  return(x)
}))

count_N = t(apply(count[111:249], 1, function(x) {
                  x[is.na(x)] <- median(x, na.rm = TRUE)
                  return(x)
}))


count= cbind(count_T, count_N)
```

## create metadata and design matrix  

```{r}
metadata= data.frame(sample= colnames(count),
                     condition= ifelse(grepl("Tumor", colnames(count)), "Tumor", "Normal"))

metadata$condition= factor(metadata$condition, levels = c("Normal", "Tumor"))
design= model.matrix(~ metadata$condition) %>% as.data.frame()
colnames(design)= gsub("metadata\\$condition", "",colnames(design) )
```

##  in case that we have beta value, transform it to M values
##  beta_values= data %>% dplyr::select(where(is.numeric))
##  M_values <- log2(beta_values / (1 - beta_values))
##  count=  M_values
##  Basis of limma voom variance stabilization
##  The shrunken variance for gene \( i \) is calculated as:
##    
##    $$
##    s_i^2 = \frac{d_0 s_0^2 + d_i \hat{\sigma}_i^2}{d_0 + d_i}
##  $$
## si= d0s0 + diσi / d0+di
##  These parameters control the extent to which the gene-specific variances si
##  are shrunk toward the pooled variance estimate s0
## si  is the shrunken variance for gene 
## σi (sigma i) is the raw variance estimate from the linear model for gene 
##  s0 is the prior variance (average variance across all genes)
##  d0 is the prior degrees of freedom (reflects how much information is borrowed from the prior)
##  di is the degrees of freedom associated with gene i
## Genes with noisy estimates (low di) have their variances more heavily shrunk towards the prior 
##  Use voom to transform the raw counts and estimate mean-variance trend

```{r}
v <- voom(count, design)
```

##  Fit the linear model

```{r}
fit <- lmFit(v, design)
```

##  Apply empirical Bayes moderation with trend adjustment

```{r}
fit <- eBayes(fit, trend = TRUE)
```

## get coefficients

```{r}
coef(fit) %>% head()
```

##  Get the top differentially expressed genes

```{r}
topTable <- topTable(fit, coef = 2,number = Inf)
sig= topTable[topTable$adj.P.Val < .01 & abs(topTable$logFC) > log2(1) ,]

write.csv(sig, "output/DMGs.csv")
```

##  Get the results for a specific contrast
##  coef(fit) %>% head()
##  contrast <- makeContrasts(Tumor - Normal, levels = design)
##  fit2 <- contrasts.fit(fit, contrast)
##  fit2 <- eBayes(fit2)
##  
##  #' ##  Get the top differentially expressed genes
##  topTable <- topTable(fit2, number = Inf)
## heatmap

```{r}
count= as.data.frame(count)
sig= topTable[topTable$adj.P.Val < .01 & abs(topTable$logFC) > log2(1.2) ,]
heat_data = count[row.names(count) %in% row.names(sig),] 
heat_data= scale(heat_data)


ta <- HeatmapAnnotation(
  Condition = metadata[, c("condition")],
  col = list(
    Condition = c("Tumor" = "#9400D3" , "Normal" = "darkolivegreen")
    
  ),
  annotation_height = unit(10, "mm")
)

heatmap  <- Heatmap(
  matrix = as.matrix(heat_data),
  name = "Normalized Express.data",
  col = colorRamp2(c(-3,0, 3), c( "grey36","lightyellow2" ,"#FFA500CC")),
  show_row_names = TRUE,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_column_names = FALSE,
  top_annotation  = ta,
  heatmap_legend_param = list(
    title_gp = gpar(fontsize = 12),  #' ##  Title text size
    labels_gp = gpar(fontsize = 12), #' ##  Labels text size
    legend_height = unit(9, "cm"),   #' ##  Height of the legend
    legend_width = unit(1.5, "cm")   #' ##  Width of the legend
  )
) 

print(heatmap)
```

## print(heatmap)

```{r}
png("output/DMGs_heatmap.png",width = 6000, height = 7000, res = 600)
draw(heatmap, annotation_legend_side =  "right")
dev.off()
```

